Left-linear rules in adhesive categories

### 2024-02-06 icalp/fusione-pratendsec2.tex
bozza altra versione intro

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndii}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofii{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndii}]\phantomsection\label{proof:prAtEndii}We can start considering the cube below and noticing that $n$, being the pushout of $m\in \mathcal {M}$, belongs to $\mathcal {A}$ too. \[\xymatrix @C=13pt@R=13pt{&A\ar [dd]|\hole _(.65){\id {A}}\ar [rr]^{g} \ar [dl]_{\id {A}} && B \ar [dd]^{\id {B}} \ar [dl]_{\id {B}} \\ A \ar [dd]_{m}\ar [rr]^(.65){g} & & B \ar [dd]_(.3){n}\\&A\ar [rr]|\hole ^(.65){g} \ar [dl]^{m} && B \ar [dl]^{n} \\C \ar [rr]_{f} & & D}\] By construction the top face of the cube is a pushout and the back one a pullback. The left face is a pullback because $m$ is mono. Thus the $\mathcal {A}$-Van Kampen property yields that the front and the right faces are pullbacks.\end{proof}

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndiii}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofiii{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndiii}]\phantomsection\label{proof:prAtEndiii}$(\Rightarrow )$ By hypothesis there exists $k:D'\to B$ such that $n\circ k = d$. By \Cref {prop:pbpo}, the bottom face of the cube is a pullback. Thus there exists a unique $h:C'\to A$ as in the diagram below, implying the thesis. \[\xymatrix @C=40pt{C' \ar [d]^{h}\ar [r]^{f'} \ar @/_.5cm/[dd]_{c}& D' \ar @/^.5cm/[dd]^{d}\ar [d]_{k}\\ A \ar [d]^{m} \ar [r]^{g} & B \ar [d]_{n} \\ C \ar [r]_f & D}\] \par \smallskip \noindent $(\Leftarrow )$ Let $h:C\to A$ be such that $c=m\circ h$. By the $\mathcal {A}$-Van Kampen property the top face of the given cube is a pushout. Thus the dotted $k:D'\to B$ in the following diagram exists. \[\xymatrix @C=40pt{A' \ar @/_.5cm/[dd]_{a} \ar [d]^{m'} \ar [r]^{g'} & B' \ar [d]_{n'} \ar @/^.5cm/[dd]^{b} \\ C' \ar [d]^{h} \ar [r]_{f'} & D' \ar @{.>}[d]_{k}\\ A \ar [r]_g& B }\] Moreover, by construction we have \begin {align*} n\circ k \circ n'&= n\circ b=d\circ n'\\ n\circ k \circ f' &= n\circ g\circ h=f\circ m\circ h=f\circ c= d\circ f' \end {align*} We can therefore conclude that $n\circ k =d$.\end{proof}

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndiv}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofiv{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndiv}]\phantomsection\label{proof:prAtEndiv}We can begin noticing that $g$, being the pullback of $k$, is mono and in $\mathcal {A}$ too. Thus we can build the cube below, in which all the vertical faces are pullbacks, entailing that all the vertical arrows are in $\mathcal {A}$. \[\xymatrix @C=20pt@R=10pt{ & &X\ar [dl]_{f} \ar [ddd]^(.333333){\id {X}}|(.666666)\hole \ar [rrr]^{a} &&&A \ar [ddd]^{\id {A}} \ar [dl]^{h}\\& Y\ar [dl]_{\id {Y}} \ar [ddd]|(.333333)\hole _{\id {Y}} &&&B \ar [ddd]^{\id {B}} \ar [dl]_{\id {B}} \\Y \ar [ddd]_{g} \ar [rrr]^{b}&&& B\ar [ddd]^{k}\\&&X \ar [rrr]|(.34)\hole ^{a}|(.67)\hole \ar [dl]_{f}&&& A \ar [dl]^{h}\\ & Y \ar [dl]_{g}&&& B \ar [dl]^{k}\\ Z\ar [rrr]_{c} &&& C}\] By hypothesis the face is an $\mathcal {A}$-stable pushout and so its top one is a pushout. Using \Cref {lem:po1} we can conclude that the right half of the rectangle with which we have started is a pushout too.\end{proof}

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndv}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofv{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndv}]\phantomsection\label{proof:prAtEndv}Take the diagram \[\xymatrix {X \ar [d]_{\id {X}}\ar [r]^{f}& Y \ar [r]^{\id {Y}} \ar [d]_{\id {Y}}& Y \ar [d]^{m}\\ X \ar [r]_{f}& Y \ar [r]_{m} & Z}\] Since $m$ is mono the right square is a pullback, while the left square is a pullback by construction. By \Cref {lem:pb1} the whole rectangle is a pullback and the thesis follows.\end{proof}

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndvi}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofvi{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndvi}]\phantomsection\label{proof:prAtEndvi}Let $m$ be an element of $\mathcal {M}$ and consider its pushout along itself. \[\xymatrix {X\ar @{>->}[r]^m \ar @{>->}[d]_m& Y\ar @{>->}[d]^{f}\\Y \ar @{>->}[r]_g & Z}\] By \Cref {prop:pbpoad} this square is a pullback, proving that $m$ is the equalizer of the arrows $f,g\colon Y\rightrightarrows Z$.\end{proof}

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndvii}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofvii{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndvii}]\phantomsection\label{proof:prAtEndvii}\begin {enumerate} \item By \Cref {prop:stab}, it follows that both squares are pushouts, thus the thesis follows from \Cref {prop:pbpoad}. \item By hypothesis, $g$ is the pullback of an arrow in $\mathcal {M}$, thus it belongs to it. But then $g\circ f\in \mathcal {M}$ too and the whole rectangle is a $\mathcal {M}$-pushout. Therefore, by \Cref {prop:pbpoad} a pullback, so that its left half is a pullback too, by \Cref {prop:pbpo}. Moreover $k\circ h$ is in $\mathcal {M}$ as the pushout of $g\circ f$ and, by \Cref {cor:deco}, we also know that $h\in \mathcal {M}$. \par Using \Cref {lem:po1}, it is enough to show that the left half of the original rectangle is a pushout. We can build the following cube: \[\xymatrix @C=20pt@R=10pt{ & &X\ar @{>->}[dl]_{f} \ar [ddd]^(.333333){\id {X}}|(.666666)\hole \ar [rrr]^{a} &&&A \ar [ddd]^{\id {A}} \ar @{>->}[dl]^{h}\\& Y\ar [dl]_{\id {Y}} \ar [ddd]|(.333333)\hole _{\id {Y}} &&&B \ar [ddd]^{\id {B}} \ar [dl]_{\id {B}} \\Y \ar @{>->}[ddd]_{g} \ar [rrr]^{b}&&& B\ar @{>->}[ddd]^{k}\\&&X \ar [rrr]|(.34)\hole ^{a}|(.67)\hole \ar @{>->}[dl]_{f}&&& A \ar @{>->}[dl]^{h}\\ & Y \ar [rrr]|(.67)\hole ^{b} \ar @{>->}[dl]_{g}&&& B \ar @{>->}[dl]^{k}\\ Z\ar [rrr]_{c} &&& C}\] Its vertical faces are all pullbacks and all the vertical arrows are in $\mathcal {M}$, hence the top face is a pushout and we can conclude. \qedhere \end {enumerate}\end{proof}

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndviii}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofviii{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndviii}]\phantomsection\label{proof:prAtEndviii}We start pulling back $n$ along $k$ and then pulling back the resulting arrow along $h$, getting the following two pullbacks square: \[\xymatrix {B \ar @{>->}[d]_{b}\ar @{>->}[r]^{s}& N \ar @{>->}[d]^{n} & A \ar [r]^{r} \ar @{>->}[d]_{a}& B \ar @{>->}[d]^{b}\\Q\ar @{>->}[r]_{k}&Y &M \ar [r]_h&Q\\}\] Notice that \[ f\circ m\circ a=k\circ h\circ a=k\circ b\circ r=n\circ s\circ r\] \par Thus there exists a unique $t\colon A\to P$ as in the diagram \[\xymatrix {A \ar @{>->}[d]_{a} \ar @{.>}[dr]^{t}\ar [r]^{r} &B \ar @{>->}@/^.2cm/[dr]^{s} \\ M \ar @{>->}@/_.2cm/[dr]_{m}&P\ar [r]^{p_1} \ar @{>->}[d]_{p_2}& N \ar @{>->}[d]^{n}\\ &X \ar [r]_{f} & Y}\] We can then consider the cube below, in which the front, right and back faces are pullbacks. \[\xymatrix @C=15pt@R=15pt{&A\ar @{>->}[dd]|\hole _(.65){a}\ar [rr]^{r} \ar [dl]_{t} && B \ar @{>->}[dd]^{b} \ar [dl]_{s} \\ P \ar @{>->}[dd]_{p_2}\ar [rr]^(.7){p_1} & & N \ar @{>->}[dd]_(.3){n}\\&M\ar [rr]|\hole ^(.65){h} \ar @{>->}[dl]^{m} && Q \ar @{>->}[dl]^{k} \\X \ar [rr]_{f} & & Y}\] We can notice that the left square is a pullback too. To see this it is enough to apply \Cref {lem:pb1} to the diagram \[\xymatrix {A \ar @/^.4cm/[rr]^{s\circ r}\ar @{>->}[d]_{a} \ar [r]_{t}& \ar [r]_{p_1} P \ar @{>->}[d]^{p_2}& N \ar [d]^{c}\\ M \ar @/_.4cm/[rr]_{k\circ h}\ar @{>->}[r]^{m}& X \ar [r]^{f}& Y}\] \par The thesis now follows immediately from \Cref {lem:varie}. \qedhere \end{proof}

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndix}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofix{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndix}]\phantomsection\label{proof:prAtEndix}Consider the two pushout squares \[\xymatrix {M \ar [r]^{h_1}\ar @{>->}[d]_{m}& W_1 \ar @{>->}[d]^{k_1} & M \ar @{>->}[d]_{m} \ar [r]^{h_1}& W_2 \ar @{>->}[d]^{k_2}\\Y\ar [r]_{f}&Z &Y \ar [r]_{f}&Z}\] Since $\mathcal {M}$ is in $\mathcal {M}$, \cref {prop:pbpoad} guarantees that they are both pullbacks. Since $m\leq m$, by \Cref {lem:radj} we get that now entails that $k_1\leq k_2$ and $k_2\leq k_1$. Thus there exists an isomorphism $\phi \colon W_1\to W_2$ such that $k_1=k_2\circ \phi $. To see that $h_2=\phi \circ h_1$, we can compute: \[ k_2\circ \phi \circ h_1 = k_1\circ h_1= n\circ m= k_2\circ h_2\] \par The claim now follows since $k_2$ is a monomorphism.\end{proof}

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndx}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofx{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndx}]\phantomsection\label{proof:prAtEndx}By construction, the pairs $(k, f)$ and $(k', f')$ are pushout complements of $l$ and $n$. Thus, the existence of the isomorphism $t\colon D\to D'$ follows from \Cref {lem:pocomp}. Now, computing we have \[ g'\circ t \circ k= g' \circ k'=h'\circ r\] Hence, we have the wanted $s\colon H\to H'$. To see that $s$ is an isomorphism, consider the diagram \[\xymatrix {K \ar @/^.4cm/[rr]^{k'}\ar [d]_{r} \ar [r]_{k}& \ar [r]_{t} D \ar [d]^{g}& D' \ar [d]^{g'}\\ R \ar @/_.4cm/[rr]_{h'} \ar [r]^{h}& H \ar [r]^{s}& H'}\] By hypothesis the whole rectangle and its left half are pushouts, therefore, by \Cref {lem:po1} its right square is a pushout too. The claim now follows from the fact that the pushout of an isomorphism is an isomorphism.\end{proof}

>>>>

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndii}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofii{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndii}]\phantomsection\label{proof:prAtEndii}Both pairs $(k, f)$ and $(k', f')$ are pushout complements of $l$ and $n$, thus, \Cref {lem:pocomp} yields an isomorphism $\phi _D\colon D\to D'$. Computing we have \[ g'\circ \phi _D \circ k= g' \circ k'=h'\circ r\] \par \noindent \parbox {3cm}{ $\xymatrix {K \ar @/^.4cm/[rr]^{k'}\ar [d]_{r} \ar [r]_{k}& \ar [r]_{\phi _D} D \ar [d]^{g}& D' \ar [d]^{g'}\\ R \ar @/_.4cm/[rr]_{h'} \ar [r]^{h}& H \ar [r]^{\phi _H}& H'}$} \hfill \parbox {10cm}{ \hspace {15pt}Hence, we have another $\phi _H\colon H\to H'$. To see that $\phi _H$ is an isomorphism, consider the diagram aside. By hypothesis the whole rectangle and its left half are pushouts, therefore, by \Cref {lem:po1} its right square is a pushout too. The claim now follows from the fact that the pushout of an isomorphism is an isomorphism. \qedhere }\end{proof}

### 2024-02-06 icalp/fusione-pratendsec3.tex
bozza altra versione intro

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndxi}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofxi{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndxi}]\phantomsection\label{proof:prAtEndxi}The first point of \Cref {def:switc} is obviously satisfied. The second one follows at once since \[m_{\der {E},0}= f_0\circ i_1 \qquad h_{\der {E},1}= g_{1}\circ i_0\] Finally, the third point holds because we already know that \[m_{\der {E},0}= f_0\circ i_1 \qquad h_{\der {E},1}= g_{1}\circ i_0\]\end{proof}

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndxii}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofxii{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndxii}]\phantomsection\label{proof:prAtEndxii}aaa\end{proof}

>>>>

\makeatletter\Hy@SaveLastskip\label{proofsection:prAtEndiii}\ifdefined\pratend@current@sectionlike@label\immediate\write\@auxout{\string\gdef\string\pratend@section@for@proofiii{\pratend@current@sectionlike@label}}\fi\Hy@RestoreLastskip\makeatother\begin{proof}[Proof of \pratendRef{thm:prAtEndiii}]\phantomsection\label{proof:prAtEndiii}Let us begin considering the diagram built in \Cref {rem:deco}: \[\xymatrix {&&R_0 \ar @/_1cm/[ddrr]_(.35){i_0}|(.7)\hole \ar [dr]^{h_0}&& L_1\ar @/^1cm/[ddll]^(.35){i_1} \ar [dl]_{m_1}\\&K_0\ar [dr]^{k_0}\ar @{>->}[dl]_{l_0} \ar @/_.8cm/[ddrr]^(.65){u_0}\ar [ur]^{r_0}&& G_1 && K_1 \ar @/^.8cm/[ddll]_(.65){u_1}\ar [dl]_{k_1}\ar @{>->}[lu]_{l_1} \ar [dr]^{r_1}\\L_0 \ar [dr]^{m_0} \ar @{>}@/_.8cm/[ddrr]_{j_1}&& D_0 \ar @{>->}[dl]|(.42)\hole _(.64){f_0}\ar [ur]|(.48)\hole ^(.7){g_1}&&D_1 \ar [dr]|(.42)\hole ^(.65){g_1} \ar @{>->}[ul]|(.48)\hole _(.7){f_1}&&R_1\ar @/^.8cm/[ddll]^{j_0}\ar [dl]^{h_1}\\&G_0 &&P\ar [dr]^{q_1} \ar @{>->}[dl]_{q_0}\ar [ur]^(.4){p_1}\ar @{>->}[ul]_(.4){p_0}&&G_2\\&&Q_0 \ar @{>->}@{>->}[ul]_{a_0} \ar @{.>}[dr]_{b_0} &&Q_1\ar [ur]^{b_1} \ar @{>.>}[dl]^{a_1} \\ &&&H_1 }\] Since $q_0\colon P\to Q_0$ is in $\mathcal {M}$, we can take its pushout along $q_1$ to get the dotted arrows $a_1\colon Q_1\to H_1$ and $q_0\colon Q_0\to H_1$. But then, since the composite of two pushout squares is a pushout, the diagram below is a derivation. \[\xymatrix @C=15pt{L_1 \ar [d]_{f_0 \circ i_1}&& K_1 \ar [d]_{q_0\circ u_1}\ar @{>->}[ll]_{l_1} \ar [r]^{r_1} & R_1 \ar @/^.35cm/[drrr]|(.3)\hole _(.4){j_0} \ar [dr]|(.3)\hole _{a_1\circ j_0} && L_0 \ar @/_.35cm/[dlll]^(.4){j_1} \ar [dl]|(.3)\hole ^{b_0\circ j_1}& K_1 \ar [d]^{q_1\circ u_0}\ar @{>->}[l]_{l_0} \ar [rr]^{r_0} && R_0 \ar [d]^{g_1\circ i_0} \\G_0 && \ar @{>->}[ll]^{a_0} D_0 \ar [rr]_{b_0}&& G_1 && \ar @{>->}[ll]^{a_1} D_1 \ar [rr]_{b_1}&& G_2}\] It is immediate now to see that it is a switch of $\dder {D}_0$ and $\dder {D}_1$ along $(i_0, i_1)$.\end{proof}

### 2024-02-06 icalp/fusione.tex
bozza altra versione intro

\todo{A VERY NICE INTRODUCTION}

SOME IDEAS (BADLY WRITTEN, JUST TO START TO PUT IDEAS IN ORDER)

Rewriting systems over graphs are used in uncountably many setting for
representing the dynamics of software systems: the graph serves to represent
the entities and their relations within states , while rewriting steps,
modifying the graph, model computational steps that results in
changes to the system state.

@@

In some situations rewriting with linear rules are not sufficiently
expressive for capturing the dynamic of a system. This is the case
when a computational step is required not only to consume, preserve
and generated, but also ``merge'' parts of the state. This happens,
e.g., in nominal calculi where, as a result of name passing, the
received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimizations~\cite{WNW:egg}. Rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

\todo{P: something about string diagrams?}

>>>>

%\todo{A VERY NICE INTRODUCTION}
%
%SOME IDEAS (BADLY WRITTEN, JUST TO START TO PUT IDEAS IN ORDER)

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, this equivalence 
captures when steps are causally unrelated and thus could be executed in any 
order, and possibly concurrently. 
%
Two of the seminal contribution to this line of research have been
Mazurkiewicz's traces~\cite{xxx} and Winskel's event structures~\cite{xxx}
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules which have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
As noteworthy examples, we have 
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that results 
in changes to the system state. The DPO (double-pushout) approach~\cite{xxx} is
nowadays considered the standard one for these structures, mostly thanks to its 
flexibility: the rules allow to specify that, 
in a rewriting step, some entities in the current state are removed, some others are 
needed for the rewriting step to occur but remain unaltered and some new entities 
can be generated. Concerning concurrency, the chosen notion is shift equivalence,
and independence is formally captured by the so-called Church-Rosser 
theorem of DPO rewriting~\cite{CMREHL:AAGT}.

A noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to devise 
the foundations of a rewriting theory which can be subsequently instantiated 
to the context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting. In fact, the framework is
so general that it allowed to recast in the DPO approach many additional formalisms,
ranging from xxx to string diagrams~\cite{bonchi2022string}.

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required only to consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagram mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

=========DA QUI SUPERSEDED\\
Rewriting systems over graphs are used in uncountably many setting for
representing the dynamics of software systems: the graph serves to represent
the entities and their relations within states , while rewriting steps,
modifying the graph, model computational steps that results in
changes to the system state.

@@

In some situations rewriting with linear rules are not sufficiently
expressive for capturing the dynamic of a system. This is the case
when a computational step is required not only to consume, preserve
and generated, but also ``merge'' parts of the state. This happens,
e.g., in nominal calculi where, as a result of name passing, the
received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}. Rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}\\
=========
